Mendelian randomization with genome-wide association studies data

### 2022-04-04 reply-to-reviewers.Rmd

---
title: "Reply to reviewers"
author: "Fred Boehm"
date: "4/4/2022"
output: html_document
---

Editor and Reviewer comments:

With pleasure I read this manuscript that compares methods for Mendelian randomization (MR). I think the authors did a lot of work in making such an extensive review. However, I do have some comments:

-I think the current manuscript is somewhere in between guiding  people who are new to MR and giving a complete overview of all methods. I think adding a standard MR dag would be beneficial to the first type of reader. This could be part of Fig 3 perhaps.

-On page 13, the authors write 'This can happen, for instance, by using a noncausal SNP that is in high LD with a causal SNP. Second, because many traits are polygenic or omnigenic, multiple causal SNPs may be
in high LD together. LD clumping is likely to omit one or more of the causal SNPs, which candiminish statistical power in MR analysis.' (p13). I think this is not too much of a concern, as long as there is strong correlation between these SNPs, because the GWAS estimate of a single variant also takes into account it's LD structure with other SNPs. Hence, it is not really only the effect of that specific SNP that is measured and hence not much information should be lost due to clumping.

-The current section 'Applications of MR in GWAS' seems quite random why these specific studies were chosen. To me as a reader I wonder why are these specific studies chosen.

-I think the manuscript could give more guidance to readers on what methods should be used or not. This could easily extend Figure 4. Currently it is unclear how different these methods really are and when they will give different results.

-I think the manuscript would improve if it would say something on future prospects (this is somewhat done in the section summary and outlook, but is currently a bit underwhelming). E.g. what are currently open challenges?

Minor remarks:

-Please beware that Mendelian randomization should be written with capital M (e.g. caption Figure 1).

-mathematical symbols sometimes difficult to read (\hat{} command often overlap)

-I find the graphical abstract unclear.

### 2022-04-26 reply-to-review1/mr-review.Rmd
copied chunks of mr-review.txt into mr-review.Rmd

Complex trait genetics studies in humans have yielded thousands of new associations for  
diverse traits. The next step in understanding biology of health and disease is to
identify causal relationships among complex traits. Two-sample Mendelian randomization studies 
in genome-wide association studies can uncover causal relationships 
between complex traits. By leveraging shared SNP genotypes as instrumental variables, or proxies, for 
the exposure complex trait, investigators can tease out causal effects from observational data, provided
that necessary assumptions are satisfied. We discuss below the growth of genome-wide association studies 
in parallel with development of Mendelian randomization methods. We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.

@@

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012]. Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of 
human health and disease.

@@

A number of excellent MR reviews have appeared in the literature [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022]. Our review doesn't 
aim to replace them, but seeks to complement their findings. 
We present recent methods developments for an audience of statisticians and epidemiologists. We discuss in detail methods advances before briefly summarizing 
applications and directions for future research. The value of MR analysis with 
GWAS data will continue to grow in parallel with statistical methods advances.

Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
publication of the early GWAS (Figure \@ref(fig:article-counts)). Many studies use the "two-sample" MR 
design, in which one cohort of subjects has measurements for the exposure, while a second cohort has 
measurements for a second trait [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in 
both sets of subjects.

Recent methods developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These strategies are especially pertinent now that there is 
an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
ability to use published data has contributed to a rapid increase in the number of
two-sample MR methods for GWAS summary statistics (Figure \@ref(fig:article-counts)).

>>>>

Genome-wide association studies have yielded thousands of associations for many common diseases and disease-related complex traits. The identified associations made it possible to identify the causal risk factors underlying diseases and investigate the causal relationships among complex traits through Mendelian randomization. Mendelian randomization is a form of instrumental variable analysis that uses SNP associations from genome-wide association studies as instruments to study and uncover causal relationships between complex traits. By leveraging SNP genotypes as instrumental variables, or proxies, for the exposure complex trait, investigators can tease out causal effects from observational data, provided that necessary assumptions are satisfied. We discuss below the development of Mendelian randomization methods in parallel with the growth of genome-wide association studies. We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods with relaxed causality assumptions and that this area continues to offer opportunities for robust biological discoveries.

@@

Great interest has developed in inferring causal relationships between complex traits, i.e., traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to enhancing our understanding of the biology of health and disease. Inferring causal relationship in genetics is often carried out through Mendelian randomization (MR) analysis. MR analysis is a form of instrumental variable analysis in which genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian]. 
MR analysis is facilitated by the development of genome-wide association studies (GWAS), which present unique opportunities for discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS interrogates millions of single nucleotide polymorphisms (SNPs) to infer which are associated with the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005]. In the nearly 17 years since publication of early studies, researchers have reported thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012]. Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of human health and disease. The association results from many large-scale GWAS are nowadays readily available, often in the form of summary statistics that include the marginal SNP p-values and/or their effect size estimates and standard errors[@buniello2019nhgri]. The identified SNP associations are used as the main input for MR analysis and thus the wide availability of GWAS summary statistics clear the way for effective MR analysis in complex genetics studies.

@@

Effective MR analyses are enabled by many MR methods developed in recent years. Methods for and uses of MR have appeared at a rapid and accelerating pace since the publication of the early GWAS (Figure 1). The overall trend of the methodology development is in the direction of increasingly sophisticated modeling of horizontal pleiotropy including both independent and correlated horizontal pleiotropy while attempting to maintain scalability and computational efficiency in the presence of multiple correlated SNPs. Here, we present a comprehensive review on 47 MR methods, primarily in the context of GWAS summary statistics, to help practitioners to choose which MR methods to use in applied data analysis. We discuss the basics form of MR analysis, the causality assumptions in MR, and how recent MR methods are developed to ensure robust results in the presence of assumption violations. We discuss in detail methods advances before briefly summarizing applications and directions for future research. Different from existing MR reviews [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022], we discuss in detail recent methods developments that enable modeling of horizontal pleiotropy and correlated horizontal pleiotropy and place these developments in the larger context of MR analysis with GWAS data. We present these recent methods developments for an audience of both statisticians and epidemiologists. We hope our review will facilitate the further advance of MR methods and their wide application on GWAS data.

## Assumptions of Mendelian randomization

Mendelian randomization uses genetic markers in the form of SNP genotypes as the proxy (“instrument” or “instrumental variable”) for the “exposure,” a complex trait, and asks whether the “exposure” variable is causal for the “outcome” variable, which is typically a second complex trait. The exposure and outcome variables can be binary, a count, a time to event, or a continuous variable. For brevity and simplicity, we focus on continuous exposures and outcomes before considering other classes of outcome variables in a later section. Most MR studies use the “two-sample” MR design, in which one cohort of subjects has measurements for the exposure, while a second cohort has measurements for the outcome, with both cohorts sharing the same set of SNP instruments [@pierce_efficient_2013]. We focus on the two-sample MR design before considering the one-sample design and the partial two-sample design where samples are partially overlapped between the two cohorts.  
MR aims to infer the causal effect of the exposure variable on the outcome variable in observational studies. The proper causal interpretation in an MR study requires the SNP instruments to satisfy three causality assumptions. 
The first assumption states that SNP instruments are associated with the exposure. If the first assumption, about the association between the SNP instruments and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014]. Indeed, bias in causal estimates can increase with decreases in the strength of association between the SNP instrument and exposure [@burgess_avoiding_2011]. 
The second assumption requires independence, conditional on the exposure and all measured and unmeasured confounders, between the SNP instruments and the outcome. The second MR assumption, often termed the exclusion restriction assumption, can also be stated as the need for the SNP instruments to affect the outcome only through the exposure. In a causal diagram, this means that the only path from the SNP instruments to the outcome is that containing the exposure. 
The last of the three assumptions states that the SNP instruments are independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome. The third assumption ensures that the SNP instruments are independent of all confounders of the exposure-outcome relationship. 
In practice, careful evaluation of the assumptions is needed. Because assumptions 2 and 3, which involve absence of certain types of confounding, can’t be verified, practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014]. We will discuss these sensitivity analyses in more detail in a late section and consider their evaluation in MR studies [@vanderweele_methodological_2014].
The causal assumptions of MR also help one to understand the origin of the term “Mendelian randomization”. Recall Mendel’s inheritance laws [mendel1865versuche], under the assumption that alleles segregate randomly from parent to offspring, the offspring genotypes are unlikely to be associated with confounders of the exposure-outcome relationship. Additionally, reverse causation from the outcome or exposure to the genotypes is unanticipated since germ-line genotypes are fixed at conception and, thus, precede realization of other observed variables. Therefore, using SNP instruments in the instrumental variable analysis is often referred to as Mendelian randomization.

## Statistical models and methods for MR with one instrument
We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where there is a single SNP instrument and a single outcome variable. Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods that leverage multiple SNP instruments.

### Ratio of coefficients method
The ratio of coefficients method, also known as the Wald method, estimates the causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. For a continuous outcome, the causal effect estimator,  is defined in Equation (3.1).

In Equation (3.1),  and  are the slope estimates from the regressions of the outcome and exposure, respectively, on the SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with summary data. However, the Wald method doesn’t accommodate multiple SNP instruments, which limits its direct use in the GWAS setting as will be discussed later.

### Two-stage methods
A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous outcome, one may perform “two-stage least squares,” which involves two linear regressions. First, the exposure variable is regressed on the instrument (i.e., the SNP genotypes) (Equation (3.2)). We denote each subject’s exposure variable value as  in Equation (3.2), while  denotes an intercept term and  is the slope.  denotes the SNP instrument genotype for subject , while , the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.
The resulting fitted values for the exposure variable from Equation (3.2) are the independent variable in the second linear regression, where the outcome is the dependent variable. The causal effect estimate, then, is the coefficient  obtained from the second regression analysis (Equation (3.3)). Note also that the random errors in Equation (3.3) are assumed independent of those in Equation (3.2). Like the  in Equation (3.2), the  are assumed independent and identically distributed normal random errors.

Note that the uncertainty in the fitted values from the first regression is not considered when performing the second regression. For this reason, the variance of the coefficient estimator,  is incorrect in two-stage calculations [@burgess_review_2017;@angrist2008mostly]. This and other observations led researchers to develop likelihood-based MR methods.

### Likelihood-based methods
Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations (3.4) and (3.5)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].

Limited information maximum likelihood is sometimes called the maximum likelihood counterpart of two-stage least squares, and it yields the same causal estimate as two-stage least squares and the ratio method when used with a single SNP instrument. Additionally, the limited information maximum likelihood framework can accommodate  SNP instruments by replacing  with the sum .
One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003]. @burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation (3.6)). The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter,  represents the effect between the true means for the exposure and the outcome. In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.

For each subject , the exposure and outcome values come from a bivariate normal distribution. The mean of the exposure distribution is assumed linear in the SNP instrument, and the mean of the outcome distribution is linear in the mean exposure [@burgess_review_2017;@jones_choice_2012]. @burgess_improving_2012 demonstrated that the Bayesian model, with weak instruments, outperforms the limited information maximum likelihood model in terms of frequentist coverage levels.
### Semiparametric methods
Semiparametric instrumental variable methods, which feature parametric and nonparametric components, typically assume a parametric model connecting the exposure and outcome, but don’t impose distributional assumptions on the errors [@burgess_review_2017]. Compared to fully parametric models, semiparametric models often are more robust to model misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow [@burgess_review_2017] by discussing three semiparametric strategies, generalized method of moments, continuous updating estimator, and G-estimation of structural mean models.
Generalized method of moments can be viewed as a more flexible form of two-stage least squares that handles heteroscedastic errors and nonlinearity in the two regressions [@burgess_review_2017]. Before specifying the model, we introduce notation.  is the conditional expectation of  if we forced  to take value  for every subject [@pearl2009causality].
The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation (3.7).

The GMM estimate is the value of the vector  that satisfies Equation (3.7), where . @palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations (3.7).
Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].
We follow [@burgess_review_2017] by defining the potential outcome  as the outcome value that we would have observed had we set the exposure value  to . For example,  denotes the observed outcome had we set the exposure to zero instead of its observed value . The structural mean model for a continuous outcome is displayed in Equation (3.8).

The causal effect parameter is . [@burgess_review_2017] derives the estimating equations for , after noting that the conditional expectation,  is independent of  and reasoning that the causal effect is that value of  that yields zero covariance between  and  (Equations (3.9)).

Note that  indexes SNP instruments and ranges from  to  in Equation (3.9).

## Mendelian Randomization with multiple independent SNP instruments
While using a single SNP instrument for MR analysis can be effective, @smith_mendelian_2004 recognized that many genetic variants individually explain a small proportion of the variation in a trait, and, thus, sufficient statistical power of MR analysis with a single SNP instrument would require sample sizes in the tens of thousands [@burgess_mendelian_2013;@schatzkin2009mendelian]. @palmer_using_2012 resolved this issue by proposing use of multiple SNP instruments in MR analysis. In particular, @pierce_power_2011 reasoned that a single causal estimate derived from a collection of causal SNPs would have greater precision than an estimate derived from only one SNP [@burgess_mendelian_2013]. Recognition of complex traits’ diverse genetic architectures has thus fueled development of two-sample MR methods with multiple SNP instruments [@bowden_mendelian_2015;@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011]. 
There are two important considerations for MR analysis with multiple SNP instruments. The first consideration is how to choose a subset of the available SNPs to serve as instruments. The solution to this task differs among published methods and majority of MR methods chose a set of independent SNPs as instruments [@verbanck_detection_2018;@zhu_causal_2018]. The independent SNPs may be selected through linkage disequilibrium (LD) clumping [@purcell2007plink]. The second consideration is how to make use of GWAS summary statistics for MR analysis, as sharing of GWAS summary statistics was encouraged and became a common practice [@nature2012asking]. @harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised a strategy for determining a single causal effect from summary statistics for a collection of independent SNPs.  Below, we discuss a number of MR methods that make use of multiple independent SNP instruments.
One approach to integrating multiple SNP instruments into a MR framework is through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 calculate an unweighted score as the number of exposure-increasing alleles in the subject’s genotypes. They also calculate a weighted allele score by using the exposure effects as weights. For example, a subject with  copies of exposure-increasing alleles for SNP  has an unweighted allele score  and a weighted allele score . @burgess_use_2013 found that when weights are obtained from external data or from cross-validation or jackknife approaches applied to the analysis data, the allele score functions as a single instrumental variable and greatly diminishes the bias compared to that of the two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
@kang_instrumental_2016 developed a LASSO-based method, sisVIVE, to identify invalid SNP instruments [@tibshirani_regression_1996]. sisVIVE’s advantage over earlier methods is that it doesn’t require the analyst to know which SNP instruments are valid. Instead, it requires that at least 50% of the instruments be valid. sisVIVE outperforms two-stage least squares in many ways and performs similar to oracle two-stage least squares. Simulations and data analysis results reveal that sisVIVE is robust to possibly invalid instruments.
@windmeijer_use_2019, building on the research from @kang_instrumental_2016, implemented an adaptive LASSO-based estimator after recognizing that sisVIVE misclassifies valid SNP instruments as invalid when the invalid SNP instruments have strong effects on the exposure. Consistent selection of invalid SNP instruments, they found, depends on SNP instrument correlations. To address this issue, @windmeijer_use_2019 proposed a median estimator with consistency that doesn’t depend on SNP instrument-exposure association strength or the SNP instrument correlation structure. They then applied methods from @zou_adaptive_2006 to achieve a consistent estimator with the same asymptotic distribution as the oracle two-stages least squares. One important limitation of this thread of research is that both @kang_instrumental_2016 and @windmeijer_use_2019 require individual-level data.
@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods. Central to their work is the proposal to create a new instrumental variable as a linear combination of genotype data from a collection of SNP instruments. They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically
pleiotropic traits. For a collection of  SNPs, they choose the -vector that meets these criteria and maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a  penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. @jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are unbiased in their simulation settings.
@spiller_detecting_2018 and @spiller_interaction-based_2020 present a method, MR-GxE, that exploits gene by environment interactions to detect and to adjust for bias due to horizontal pleiotropy. @spiller_interaction-based_2020 treat a SNP-covariate interaction as an instrumental variable, and, by so doing, they impose assumptions on that interaction. Together, these assumptions require that MR-GxE assume that horizontal pleiotropy effects are constant across the study cohort. The authors devise a three-step inferential procedure for two-sample MR with GWAS summary statistics. First, they estimate the SNP-exposure and SNP-outcome associations at a range of covariate values. Second, they regress the SNP-outcome estimated associations on the SNP-exposure estimated associations. Third, they treat the resulting slope as the causal effect estimate and the intercept as the mean horizontal pleiotropy effect.

## Methods with Summary data
The proliferation of publicly available GWAS data accelerates the development of MR methods with multiple SNP instruments that make use of GWAS summary statistics for model fitting. Our timeline highlights some of the many multi-instrument MR methods with GWAS summary statistics, with an increasing density of methods over time in the last three years (Figure 2). These methods include the inverse variance weighted MR [61], MR-Egger [51], weighted median estimation [68], Bayesian weighted MR [69], robust adjusted profile score [70], MRMix [71], CAUSE [72], and MRAID [73]. These methods differ in their approaches to three considerations that include instrument selection (from among the available SNPs), modeling and controlling for horizontal pleiotropy, and statistical inference procedure. We summarize some of these MR methods that use independent or correlated SNP instruments below. We will cover the remaining MR methods that model horizontal pleiotropy in a later section.

Figure 2: Two-sample MR with GWAS summary statistics methods publications density increases over time.
[61] termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
Specifically, [61] combined causal effect ratio estimates from  independent SNPs by using Equation (5.1).

[68], noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
weighted median-based estimators, building on research from [94]. Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [95]. Specifically, let  denote the  ordered ratio estimate statistic, from least to greatest. If an odd number, say , of SNP instruments is used, then the simple median estimator is defined as the  ordered ratio estimate. For an even number, say , number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, .
Due to the inefficiency of the simple median estimator, [68] define the weighted median estimator. To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator. The weighted median estimator, then is the median of a distribution defined by having  as the  percentile, where  is the weight assigned to the  ordered ratio estimate. In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
[68] also studied the penalized weighted median estimator. They defined the penalized weights, , where  is the p-value resulting from the comparison of  to a  distribution [68,96]. The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [14,61–63]. In addition, LD clumping for choosing instrument SNPs can be problematic for two reasons. First, the chosen SNPs may not be causal for the outcome. This can happen, for instance, by using a noncausal SNP that is in high LD with a causal SNP. Second, because many traits are polygenic or omnigenic, multiple causal SNPs may be in high LD together. LD clumping is likely to omit one or more of the causal SNPs, which can diminish statistical power in MR analysis. In support of this is the finding that including multiple correlated SNPs in transcriptome-wide association studies improves power compared to using independent SNPs only [64–67]. Consequently, many MR methods have been recently developed to model multiple correlated instruments. 
[62] developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation. However, many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to be statistically independent. This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [61].
